---
title: "Section 1: INTRODUCTION"
author: "Pratheepa Jeganathan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: BiocStyle::html_document
---

<!-- --- -->
<!-- title: "Section 1: Introduction" -->
<!-- output: rmarkdown::html_vignette -->
<!-- --- -->

<!-- vignette: > -->
<!--   %\VignetteIndexEntry{Reproducible Research Appendix} -->
<!--   %\VignetteEngine{knitr::rmarkdown} -->
<!--   %\VignetteEncoding{UTF-8} -->

This document serves as an appendix to the reproducible research for the paper on ``A Statistical Perspective on the Challenges in Molecular Microbial Biology" in The Journal of Agricultural, Biological and Environmental Statistics.

The following articles provide description of four components of microbiome data, description of an example dataset we used in the Jeganathan et al. (2021), demonstrate a few statistical tools that can overcome some of the difficulties in analyzing molecular microbial data, and introduce differential topic analysis that can facilitate inferences on latent microbial communities.

To do reproducible analysis of the research in Jeganathan et al. (2021), we suggest to install ``diffTop`` package. 

```{r include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r eval=FALSE}
remotes::install_github("PratheepaJ/diffTop")
```

Next we load the package ``diffTop``

```{r eval=FALSE}
library(diffTop)
```

Now we will walk you through the sections in Jeganathan et al. (2021) to reproduce the analysis. 
