---
title: "Section 6: VISUALIZATIONS FOR HETEROGENEOUS DATA"
author: "Pratheepa Jeganathan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: BiocStyle::html_document
---


```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
  fig.width = 12, 
  fig.height = 12, 
  fig.align = 'center',
  message = FALSE, 
  warning = FALSE)
```

```{r}
library(phyloseq)
library(tidyverse)
library(genefilter) #KOverA
library(pheatmap)
devtools::load_all()
```


```{r}
theme_set(theme_minimal())
theme_update(
  text = element_text(size = 10),
  legend.text = element_text(size = 10)
)
```

## Data

```{r}
data("psE_BARBI")
threshold <- kOverA(2, A = 25) 
psE_BARBI <- phyloseq::filter_taxa(
  psE_BARBI, 
  threshold, TRUE) 

psE_BARBI

ps <- psE_BARBI
rm(psE_BARBI)
ps <- prune_taxa(taxa_sums(ps) > 0, ps)
ps

```

### Edit specimen names

We edit specimen names and identify Asteraceae and non-Asteraceae plants.

```{r}
sam_names <- str_replace(sample_names(ps), "E106", "E-106")
sam_names <- str_replace(sam_names, "_F_filt.fastq.gz", "")
sam_names <- str_replace(sam_names, "Connor-", "E")
sample_names(ps) <- sam_names
sample_data(ps)$X <- sam_names
sample_data(ps)$unique_names <- sam_names
aster <- c("142","143","15","ST","22","40")
non_aster <- c("33", "71", "106")

paired_aster <- c("E-142-1", "E142-1", "E-142-5", "E142-5", "E-142-10", "E142-10", "E-143-2", "E143-2", "E-143-7", "E143-7", "E-15-1", "E15-1", "ST-CAZ-4-R-O", "ST-CAZ-4-R-M", "ST-SAL-22-R-O", "ST-SAL-22-R-M", "ST-TRI-10-R-O", "ST-TRI-10-R-M")
paired_non_aster <- c("E33-7", "E-33-7", "E33-8", "E-33-8", "E33-9", "E-33-9", "E71-10", "E-71-10","E71-2","E-71-2" ,"E71-3", "E-71-3", "E106-1", "E-106-1", "E106-3", "E-106-3", "E106-4", "E-106-4")
paired_specimens <- c(paired_aster, paired_non_aster)
```

```{r}
# We will use 9 colors for 9 different plants
plant_colors <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "purple")
```

## Barplots for estimates of $p_{j}=(p_{i,j}, i=1\ldots m)$ 

For this section, we consider the example data set after the contamination removal. There are 1418 taxa in 86 samples. We compute the relative abundance of ASVs in each specimen. Then, we remove ASVs corresponding to the Class with less than five ASVs. 

```{r}
ps_ra <- transform_sample_counts(
  ps, 
  function(x){x / sum(x)}
  )
```

```{r}
remov_asvs <- names(table(tax_table(ps)[, "Class"]))[which(table(tax_table(ps)[, "Class"]) <= 5)]

ps_ra <- subset_taxa(ps_ra, 
                     !(Class %in% remov_asvs) & !is.na(Class)
                     )
ps_temp <- subset_taxa(ps, 
                       !(Class %in% remov_asvs) & !is.na(Class)
                       )
ps_temp
```

We plot the distribution of relative abundance of each Phylum in specimens faceted by Class. We can use the barplots to compare difference in scale and distribution of Phylum in universal and Aster-modified pPNA types. 

```{r}
phylum_order_by_num_asv <- sort(
  table(tax_table(ps_temp)[, "Phylum"]), 
  decreasing = TRUE
  )

# most number of ASVs are from Proteobacteria
ps_proteobacteria <- subset_taxa(
  ps_temp, 
  Phylum %in% c("Proteobacteria")
  )

ps_ra_proteobacteria <- subset_taxa(
  ps_ra,
  Phylum %in% c("Proteobacteria")
  )

plot_abundance <- function(
  physeq, 
  title = "",
  Facet = "Class", 
  Color = "Phylum"){
    mphyseq <- psmelt(physeq)
    mphyseq <- subset(
      mphyseq, 
      Abundance > 0
      )
    ggplot(
      data = mphyseq,
      mapping = aes_string(
        x = "pna",
        y = "Abundance",
        color = Color, 
        fill = Color)
      ) +
      geom_violin(fill = NA) +
      geom_point(
        size = 1, 
        alpha = 0.3,
        position = position_jitter(width = 0.3)
        ) +
      facet_wrap(facets = Facet) + 
      scale_y_log10() + 
      ggtitle(title)+
      xlab("pPNA type")+
      theme(legend.position="none", 
            plot.title = element_text(hjust = 0.5))
}

```


```{r eval=FALSE}
plot_abundance(
  subset_samples(
    ps_proteobacteria, 
    Type == "aster"), 
  title = "Asteraceae") + 
  ylab("Abundance")

```

```{r fig.cap="Supplementary Figure 6: Distribution of relative abundance of four different Classes of Proteobacteria in Asteraceae plants. O and M denote universal and Asteraceae-modified pPNA types, respectively."}
plot_abundance(
  subset_samples(
    ps_ra_proteobacteria, 
    Type == "aster"),
  title = "Asteraceae") + 
  ylab("Relative abundance")


```


```{r eval=FALSE}
plot_abundance(
  subset_samples(
    ps_proteobacteria, 
    Type == "non-aster"),
  title = "non-Asteraceae"
  ) 



```


```{r fig.cap="Supplementary Figure 7: Distribution of relative abundance of four different Classes of Proteobacteria in non- Asteraceae plants. O and M denote universal and Asteraceae-modified pPNA types, respectively."}
plot_abundance(
  subset_samples(
    ps_ra_proteobacteria, 
    Type == "non-aster"), 
  title = "non-Asteraceae") + 
  ylab("Relative abundance")

```


## Heatmaps

We plot the heatmap of Anscombe's transformed values.

```{r fig.cap="Figure 3: Thirty most abundant ASVs were selected in all specimens. Taxa are labeled by Class on rows, and specimens are on the columns of the heatmap. Some specimens from Centaurea solstitialis plant have the most abundant ASVs of the Class Actinobacteria, Alphaproteobacteria, Sphingobacteria, and Gammaproteobacteria."}
data("ps_ans")
# do use asinh transformation
ps_top <- ps_ans 
    
# choose top 30 ASVs in endo specimens for heatmap
top <- names(
  sort(
    taxa_sums(
      ps_top
      ), 
    decreasing=TRUE
    )
  )[1:30]

ps_top <- prune_taxa(
  top, 
  ps_top
  )

or <- order(
  taxa_sums(ps_top), 
  decreasing=TRUE
  )[1:30]

ass <- otu_table(ps_top) %>% 
  data.frame()

colnames(ass) <- sample_names(ps_top)

tx <- tax_table(ps_top) %>% 
  data.frame()

tx <- dplyr::mutate(tx, 
                    usename = paste0(
                      "ASV_", 
                      seq(1, ntaxa(ps_top)), 
                      "_", 
                      tax_table(ps_top)[, "Class"]
                      ), 
                    rnames = taxa_names(ps_top)
                    )

ass <- ass[or, ]
rownames(ass) <- tx$usename[or]
  
df <- sample_data(ps_top) %>% 
  data.frame()

df <- df[ , c(
  "Type",
  "species_names", 
  "pna",
  "unique_names"
  )] 

colnames(df)[which(colnames(df) == "species_names")] <- "plant_type"

df <- with(
  df, 
  df[order(Type, plant_type, pna, unique_names),]
  )

ass <- ass[, df$unique_names]

df <- dplyr::select(
  df, 
  -unique_names
  )

Type <- c("royalblue3", "seagreen1")
names(Type) <- levels(df$Type)
plant_type <- plant_colors
names(plant_type) <- levels(df$plant_type)
pna <- c("orange3", "palegreen4")
names(pna) <- levels(df$pna)

ann_colors <-  list(
  Type = Type,
  plant_type = plant_type,
  pna = pna
  )

pheatmap(
  ass, 
  annotation_col = df, 
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_row = 14, 
  fontsize_col = 10,
  annotation_colors = ann_colors
  )
```



```{r include=FALSE}
rm(ann_colors, ass, df, ps, ps_proteobacteria, ps_ra, ps_ra_proteobacteria, ps_temp, ps_top, tx, aster, non_aster, or, paired_aster, paired_non_aster, paired_specimens, phylum_order_by_num_asv, plant_colors, plant_type, pna, remov_asvs, sam_names, top, Type)
```

