---
title: "Section 7: MULTIVARIATE AND NETWORK ANALYSES"
author: "Pratheepa Jeganathan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: BiocStyle::html_document
---


```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
  fig.width = 7, 
  fig.height = 4, 
  fig.align = 'center',
  message = FALSE, 
  warning = FALSE)
```

```{r}
library(phyloseq)
library(tidyverse)
library(genefilter) #KOverA
library(ggrepel) # geom_text_repel
library(randomcoloR)# distinctColorPalette(n)
library(adaptiveGPCA)
library(igraph)
library(phyloseqGraphTest)
devtools::load_all()
```


```{r}
theme_set(theme_minimal())
theme_update(
  text = element_text(size = 10),
  legend.text = element_text(size = 10)
)
```

## Data

```{r}
data("psE_BARBI")
threshold <- kOverA(2, A = 25) 
psE_BARBI <- phyloseq::filter_taxa(
  psE_BARBI, 
  threshold, TRUE) 

psE_BARBI

ps <- psE_BARBI
rm(psE_BARBI)
ps <- prune_taxa(taxa_sums(ps) > 0, ps)
ps

```

### Edit specimen names

We edit specimen names and identify Asteraceae and non-Asteraceae plants.

```{r}
sam_names <- str_replace(sample_names(ps), "E106", "E-106")
sam_names <- str_replace(sam_names, "_F_filt.fastq.gz", "")
sam_names <- str_replace(sam_names, "Connor-", "E")
sample_names(ps) <- sam_names
sample_data(ps)$X <- sam_names
sample_data(ps)$unique_names <- sam_names
aster <- c("142","143","15","ST","22","40")
non_aster <- c("33", "71", "106")

paired_aster <- c("E-142-1", "E142-1", "E-142-5", "E142-5", "E-142-10", "E142-10", "E-143-2", "E143-2", "E-143-7", "E143-7", "E-15-1", "E15-1", "ST-CAZ-4-R-O", "ST-CAZ-4-R-M", "ST-SAL-22-R-O", "ST-SAL-22-R-M", "ST-TRI-10-R-O", "ST-TRI-10-R-M")
paired_non_aster <- c("E33-7", "E-33-7", "E33-8", "E-33-8", "E33-9", "E-33-9", "E71-10", "E-71-10","E71-2","E-71-2" ,"E71-3", "E-71-3", "E106-1", "E-106-1", "E106-3", "E-106-3", "E106-4", "E-106-4")
paired_specimens <- c(paired_aster, paired_non_aster)
```


```{r}
# We will use 9 colors for 9 different plants
plant_colors <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "purple")
```

## 7.1 Ordination

### Multidimensional scaling (MDS) with weighted unifrac

```{r fig.cap="Supplementary Figure 8: Multidimensional scaling (MDS) with weighted unifrac distance on all specimens."}
ps_log <- transform_sample_counts(
  ps, 
  function(x) log(1 + x)
  )

out.wuf.log <- ordinate(
  ps_log, 
  method = "MDS", 
  distance = "wunifrac"
  )

evals <- out.wuf.log$values$Eigenvalues

mds_wuni <- plot_ordination(
  ps_log, 
  out.wuf.log, 
  color = "species_names", 
  shape = "pna"
  ) +
  labs(
    col = "Plant type"
    ) +
  coord_fixed(
    sqrt(evals[2] / evals[1])
    ) + 
  scale_colour_manual(
    values = plant_colors
    ) + 
  facet_grid(~Type) +
  geom_point(size = 3)

# adding only paired-specimen names
df_paired <- mds_wuni$data %>% 
  dplyr::filter(
    unique_names %in% paired_specimens
    )

mds_wuni + 
  geom_text_repel(
    data = df_paired, 
    aes(
      x = Axis.1, 
      y = Axis.2, 
      label = unique_names
      ), 
    color = "black"
    )


```

We see that Centaurea solstitialis specimens are outlier that were sampled in a field across three countries, while other plants are grown from seeds. We label the paired specimens in all plants. The paired specimens are relatively close to each other, except (E-143-7, E143-7), (E-142-5, E142-5), (E-71-2, E71-2), and (E-71-3, E71-3) that are in positive and negative axes of the first two principal axes. 


Fitzpatrick et al. (2018) concluded that there is no effect in within and between specimen bacterial diversity when they used universal or Aster-modified pPNA types. 

The above Figure shows that specimens from the aster plants that are paired with universal or Aster-modified pPNA types make clusters.

### MDS with BC

An MDS plot using Bray-Curtis distance between specimens.

```{r}
out.bc.log <- ordinate(
  ps_log, 
  method = "MDS", 
  distance = "bray"
  )

evals <- out.bc.log$values$Eigenvalues

mds_bc <- plot_ordination(
  ps_log, 
  out.bc.log, 
  color = "species_names", 
  shape = "pna"
  ) +
  coord_fixed(
    sqrt(evals[2] / evals[1])
    ) +
  labs(
    col = "Plant type", 
    shape = "pna"
    ) + 
  scale_colour_manual(
    values = plant_colors
    ) + 
  facet_grid(~Type) +
  geom_point(size=3) 

df_paired <- mds_bc$data %>% 
  dplyr::filter(
    unique_names %in% paired_specimens
    )

mds_bc + 
  geom_text_repel(
    data = df_paired, 
    aes(x = Axis.1, 
        y = Axis.2, 
        label = unique_names
        ), 
    color = "black"
    )

```

The ordination of specimens in MDS with BC is similar to MDS with weighted unifrac.

### DPCoA 

A DPCoA plot incorporates phylogenetic information.

The DPCoA specimens positions can be interpreted with respect to the ASVs coordinates in this display.

```{r fig.cap="Supplementary Figure 9: (A) A DPCoA plot that incorporates phylogenetic information."}
data("ps_ans")
out.dpcoa.log <- ordinate(
  ps_ans, 
  method = "DPCoA"
  )

evals <- out.dpcoa.log$eig

dpcoa_sam <- plot_ordination(
  ps_ans,
  out.dpcoa.log,
  color = "species_names",
  shape = "pna"
  ) +
  coord_fixed(
    sqrt(evals[2] / evals[1])
    ) +
  labs(
    col = "Plant type", 
    shape = "pna"
    ) + 
  scale_colour_manual(
    values = plant_colors
    ) + 
  facet_grid(~Type) +
  geom_point(size = 3) 

df_paired <- dpcoa_sam$data %>% 
  dplyr::filter(
    unique_names %in% paired_specimens
    )

dpcoa_sam + 
  geom_text_repel(
    data = df_paired,
    aes(
      x = CS1, 
      y = CS2,
      label = unique_names
      ), 
    color = "black"
    )


dpcoa_taxa <- plot_ordination(
  ps_ans, 
  out.dpcoa.log,
  type = "species", 
  color = "Class"
  ) 
  

df_taxa <- dplyr::filter(
  dpcoa_taxa$data, 
  (CS1 > 0.63 | CS1 < -0.14 | CS2 > 0.29 | CS2 < -0.13) & (
    !is.na(Class)
    )
  )

dpcoa_taxa$data <- df_taxa

dpcoa_taxa +
  coord_fixed(
    sqrt(evals[2] / evals[1])
    ) +
  geom_point(size = 2) + 
  scale_color_manual(
    values = distinctColorPalette(
      length(
        unique(
          tax_table(ps_ans)[, "Class"]
          )
        )
      )
    )

```

## 7.2 Integrating the phylogenetic tree into the analyses

### Adaptive gPCA

```{r fig.cap="Figure 4: The results from adaptive gPCA reveal Centaurea solstitialis specimens are outliers among Asteraceae plants in the left of Axis 1. Axis 2 explains the microbial variability in plant types."}
pp <- processPhyloseq(
  ps
  )
out.agpca <- adaptivegpca(
  pp$X, 
  pp$Q, 
  k = 2
  )

agPCA_samples <- ggplot(
  data.frame(
    out.agpca$U, 
    sample_data(ps)
    )
  ) +
    geom_point(
      aes(x = Axis1,
          y = Axis2, 
          color = species_names,
          shape = pna),
      size = 3
      ) +
  labs(
    x = "Axis 1", 
    y = "Axis 2",
    col = "Plant type"
    ) + 
  scale_colour_manual(
    values = plant_colors
    ) +
  facet_grid(~ Type)

agPCA_samples
```

```{r}
agPCA_asv <- ggplot(
  data.frame(
    out.agpca$QV, 
    tax_table(ps)
    )
  ) +
    geom_point(
      aes(
        x = Axis1, 
        y = Axis2, 
        color = Class
        ), 
      size = 3
      ) +
    xlab(
      "Axis 1"
      ) + 
  ylab(
    "Axis 2"
    )

main_class <- c("Bacilli", 
                "Actinobacteria", 
                "Cytophagia", 
                "Gammaproteobacteria", 
                "Betaproteobacteria", 
                "Deltaproteobacteria", 
                "Alphaproteobacteria", 
                "Opitutae")

df_taxa <- dplyr::filter(
  agPCA_asv$data, 
  Class %in% main_class
  )

agPCA_asv$data <- df_taxa

agPCA_asv <- agPCA_asv +
  scale_color_manual(
    values = distinctColorPalette(
      length(
        unique(
          tax_table(ps_ans)[, "Class"])
        )
      )
    )
agPCA_asv

```

_Centaurea solstitialis_ specimens are outliers among Asteraceae plants on the left of Axis 1. Axis 2 explains the microbial variability in plant types.

## 7.3 Correspondence analysis

```{r fig.cap="Figure 5: Correspondence analysis of all plant specimens. (A)"}
ps_ca <- ps
out.ca <- ordinate(
  ps_ca, 
  method = "CCA"
  )

evals <- out.ca$CA$eig
evals_prop <- evals/sum(evals)

ca_plot <- plot_ordination(
  ps_ca,
  out.ca,
  type = "sample",
  color = "species_names",
  shape = "pna") +
  labs(
    col = "Plant type",
    shape = "pna"
    ) + 
  scale_color_manual(
    values = plant_colors
    ) + 
  facet_grid(~Type) +
  geom_point(size=2) +
  geom_jitter()+
  coord_fixed(0.4) + 
  theme_update(
    text = element_text(size = 10),
    legend.text = element_text(size = 10),
    panel.spacing = unit(2, "lines")
)


ca_plot

```

```{r fig.cap="Figure 5: Correspondence analysis of all plant specimens. (B)"}
ca_plot_taxa <- plot_ordination(
  ps_ca,
  out.ca,
  type = "species",
  color = "Class"
  ) 

main_class <- c("Bacilli", 
                "Actinobacteria", 
                "Cytophagia", 
                "Gammaproteobacteria",
                "Betaproteobacteria",
                "Deltaproteobacteria",
                "Alphaproteobacteria", 
                "Opitutae", 
                "Sphingobacteriia")

df_taxa <- dplyr::filter(
  ca_plot_taxa$data, 
  Class %in% main_class
  )

df_taxa$Class <- factor(df_taxa$Class)

ca_plot_taxa$data <- df_taxa

ca_plot_taxa <- ca_plot_taxa +
  coord_fixed(0.4) +
  labs(col = "Class") + 
  geom_point(size=3) +
  geom_jitter() +
  scale_color_manual(
    values = distinctColorPalette(
      length(
        unique(
          main_class
          )
        )
      )
    )

ca_plot_taxa

```


## 7.4 Network analysis

### Network analysis on paired specimens

```{r}
ps <- subset_samples(
  ps, 
  unique_names %in% paired_specimens
  )
ps <- prune_taxa(
  taxa_sums(ps) > 0, 
  ps
  )
ps
```

```{r fig.cap="Supplementary Figure 10: A network created by thresholding the Jaccard dissimilarity matrix at 0.8."}
net <- make_network(
  ps, 
  max.dist = 0.8
  )

sampledata <- sample_data(ps) %>%
  data.frame()

V(net)$unique_names <- sampledata[names(V(net)), "unique_names"]
V(net)$pna <- sampledata[names(V(net)), "pna"]
V(net)$species_names <- sampledata[names(V(net)), "species_names"]

p_net <- plot_network(
  net, 
  ps, 
  color = "species_names",
  shape = "pna"
  ) + 
  labs(
    col = "Species names"
    ) + 
  scale_colour_manual(
    values = plant_colors
    )
p_net

```


### Graph-based two-sample tests


We first perform a test using an Minimum spanning tree (MST) with Jaccard dissimilarity. We want to know whether the universal or Asteraceae-modified pPNAs come from the same distribution in paired-specimens. 

Since there is a grouping in the data by specimen, we can’t simply permute all the labels, we need to maintain this nested structure – this is what the grouping argument does. Here we permute the pna labels but keep the plant type structure intact.

```{r}
set.seed(15)
gt <- graph_perm_test(
  ps, 
  sampletype = "pna", 
  grouping = "unique_names",
  distance = "jaccard", 
  type = "mst")
gt$pval
```

This test has a larger p-value, and we do not reject the null hypothesis that the paired-specimens come from the same distribution. 

This result mirrors the observations from the PCA after the truncated-rank transformation, adaptive gPCA, MDS, and DPCoA that paired-specimens have similar microbial variability.

```{r include=FALSE}
rm(df_paired, df_taxa, dpcoa_sam, dpcoa_taxa, mds_bc, mds_wuni, out.bc.log, out.dpcoa.log, out.wuf.log, ps, ps_log, agPCA_asv, agPCA_samples, out.agpca, pp, aster, evals, main_class, non_aster, paired_aster, paired_non_aster, paired_specimens, plant_colors, sam_names, ca_plot, out.ca, ps_ca, evals_prop, ca_plot_taxa, gt, net, p_net, sampledata)
```

